Normalization by Evaluation for Call-by-Push-Value and Polarized Lambda-Calculus

### 2019-02-13 fscd19/fscd19.tex
[ fscd19 ] removed definition of lookup
[ fscd19 ] reference Filinski for continuation monad

@@

and obtain a natural projection function from the semantic enironments:
  \tlookup\,(x : A \in \Gamma) & : &  \den \Gamma \todot \den A \\
  \tlookup\;\tzero     & = & \pi_2 \\
  \tlookup\;(\tsuc\,x) & = & \tlookup\;x \comp \pi_1 \\

@@

%   \tlookup & : & A \in \Gamma \to \den \Gamma \todot \den A \\

@@

@@

An alternative implementation of $\C$ is the continuation monad $\CC$
on presheaves as, for instance, employed by Ilik et al.\
\cite{ilikLeeHerbelin:apal10}
for Kripke models and NbE of classical logic.
\[
  \CC\;\J = \forall A.\ (\J \hatto \tNf\,A) \hatto \tNf\,A
\]
The answer type of this continuation monad is always $\tNF$, however,
we are polymorphic in the simple type $A$ of normal forms we produce.

@@

To add a variable of positive type $P$ to the context, we need to
break it apart until only atoms and negative bits remain.  This is
performed by maximal pattern matching, called the left-invertible
phase of focalization.
We express maximal pattern matching on $P$ as a strong functor
$\Add P$ in the category of presheaves,
mapping a presheaf $\J$ (``judgement'') to $\Add P \J$
and a presheaf morphism $f : (\J \hatto \K)\Gamma$ to
$\stmap[\Add P]_\Gamma f : \AddHyp P \J \Gamma \to \AddHyp P \K \Gamma$.
For arbitrary $\J$ and $\Gamma$, the set \fbox{$\AddHyp P \J \Gamma$} is
inductively constructed by the following rules:
\begin{gather*}
  \nru{\phyp}
      {\J\,(\cext \Gamma \patom)}
      {\AddHyp {\patom} \J \Gamma}
\qquad
  \nru{\nbranch}
      {}
      {\AddHyp {\tyzero} \J \Gamma}
\qquad
  \nru{\bbranch}
      {\AddHyp {P_1} \J \Gamma \qquad
       \AddHyp {P_2} \J \Gamma }
      {\AddHyp {P_1 + P_2} \J \Gamma}
\\[2ex]
  \nru{\nhyp}
      {\J\,(\cext \Gamma N)}
      {\AddHyp {\Thunk N} \J \Gamma}
\qquad
  \nru{\nsplit}
      {\J\,\Gamma}
      {\AddHyp {\tyone} \J \Gamma}
\qquad
  \nru{\bsplit}
      {\AddHyp {P_1} {\left(\Add{P_2}\,\J\right)} \Gamma}
      {\AddHyp {P_1 \times P_2} \J \Gamma}
\end{gather*}
Note the recursive occurrence of $\Add{P_2}$ as argument to
$\Add{P_1}$ in $\bsplit$, which makes $\Add P$ a nested datatype
\cite{bird:nested}.  Agda supports such nested inductive types; but
note that $\Add{P}$ is uncontroversial, since it could also be defined
by recursion on $P$.   It is tempting to name $\bsplit$
``$\tjoin$'' and $\nsplit$ ``$\treturn$''
since $\Add P$ is a graded monad on the monoid $(1,\times)$
of product types;
however, this coincidence shall not matter for our further considerations.

>>>>

The idea behind \emph{normalization by evaluation} (NbE) is to utilize a
standard interpreter, usually evaluating closed terms, to compute the
normal form of an open term.

@@

and obtain a natural projection function
$\tlookup\,(x : A \in \Gamma) : \den \Gamma \todot \den A$
from the semantic enironments.

@@

%   \tlookup\,(x : A \in \Gamma) & : &  \den \Gamma \todot \den A \\

@@

% % \begin{array}{lcl}
% %   \tlookup & : & A \in \Gamma \to \den \Gamma \todot \den A \\
% %   \tlookup\;\tzero     & = & \pi_2 \\
% %   \tlookup\;(\tsuc\,x) & = & \tlookup\;x \comp \pi_1 \\
% % \end{array}

@@

As already observed by Filinski
\cite[Section~5.4]{filinski:semaccounttdpe}
\cite[Section~3.2]{filinski:tlca01},
normalization by evaluation can be carried out in the continuation
monad.
In our setting, we use a continuation monad $\CC$ on presheaves
defined as
% An alternative implementation of $\C$ is the continuation monad $\CC$
% on presheaves as, for instance, employed by Ilik et al.\
% \cite{ilikLeeHerbelin:apal10}
% for Kripke models and NbE of classical logic.
\[
  \CC\;\J = \forall A.\ (\J \hatto \tNf\,A) \hatto \tNf\,A
  .
\]
The answer type of this continuation monad is always $\tNF$, however,
we are polymorphic in the simple type $A$ of normal forms we produce.%

@@

To add a variable of positive type $P$ to the context, we need to
break it apart until only atoms and negative bits remain.  This is
performed by maximal pattern matching, called the left-invertible
phase of focalization.%
\footnote{Filinski \cite[Section~4]{filinski:tlca01} achieves maximal pattern
matching through an additional, ordered context $\Theta$ for positive
variables which are eagerly split.}
We express maximal pattern matching on $P$ as a strong functor
$\Add P$ in the category of presheaves,
mapping a presheaf $\J$ (``judgement'') to $\Add P \J$
and a presheaf morphism $f : (\J \hatto \K)\Gamma$ to
$\stmap[\Add P]_\Gamma f : \AddHyp P \J \Gamma \to \AddHyp P \K \Gamma$.
For arbitrary $\J$ and $\Gamma$, the set \fbox{$\AddHyp P \J \Gamma$} is
inductively constructed by the following rules:
\begin{gather*}
  \nru{\phyp}
      {\J\,(\cext \Gamma \patom)}
      {\AddHyp {\patom} \J \Gamma}
\qquad
  \nru{\nbranch}
      {}
      {\AddHyp {\tyzero} \J \Gamma}
\qquad
  \nru{\bbranch}
      {\AddHyp {P_1} \J \Gamma \qquad
       \AddHyp {P_2} \J \Gamma }
      {\AddHyp {P_1 + P_2} \J \Gamma}
\\[2ex]
  \nru{\nhyp}
      {\J\,(\cext \Gamma N)}
      {\AddHyp {\Thunk N} \J \Gamma}
\qquad
  \nru{\nsplit}
      {\J\,\Gamma}
      {\AddHyp {\tyone} \J \Gamma}
\qquad
  \nru{\bsplit}
      {\AddHyp {P_1} {\left(\Add{P_2}\,\J\right)} \Gamma}
      {\AddHyp {P_1 \times P_2} \J \Gamma}
\end{gather*}
Note the recursive occurrence of $\Add{P_2}$ as argument to
$\Add{P_1}$ in $\bsplit$, which makes $\Add P$ a nested datatype
\cite{bird:nested}.  Agda supports such nested inductive types; but
note that $\Add{P}$ is uncontroversial, since it could also be defined
by recursion on $P$.   It is tempting to name $\bsplit$
``$\tjoin$'' and $\nsplit$ ``$\treturn$''
since $\Add P$ is a graded monad on the monoid $(1,\times)$
of product types;
however, this coincidence shall not matter for our further considerations.
